An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-01 main.tex
Dagens text

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}

@@

It is possible to compute the Parikh image of any context-free language using
the method described in~\cite{generate-parikh-image}. Later improvements have
produced a construction taking at most linear time to
produce~\cite{muscholl-linear}. However, the resulting existentially quantified
clauses are costly to eliminate, in practice making many real-world problems
intractable.

In addition to the costly elimination of the existentially quantified variables,
another problem arises for the common case of solving length constraints in
string solvers. Conjunctions of string constraints lead to the computation of
Parikh images of products of regular languages. Applying the approach
in~\cite{generate-parikh-image} would require the computation of the product
before its Parikh image. In several instances we have observed while solving
real-world string constraints, the computation of the product of automata
exhausts the memory of any machine due to the exponential blow-up in size of the
product, quickly becoming intractable as the number of automata in the product
increases.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image. This enables us to let both calculations inform each
other, eliminating unnecessary work, and pruning the size of the partial
products considered in the computation for a smaller memory footprint. Moreover,
the method can be used iteratively to tackle smaller chunks of the product
incrementally, thereby avoiding running out of memory. We implement this
calculus, along with the approximate method of~\cite{approximate-parikh}, its
fall-back variant adapted from~\cite{generate-parikh-image}, and an adapter for
the \Nuxmv{} model checker, in a tool that we call \Catra{}. \Catra{} allows
solving constraints over groups of synchronised (products of) Parikh automata,
that is (finite-alphabet) symbolic automata with transitions that increment and
decrement integer registers and that allows checking linear constraints on these
registers after running the automata. Parikh automata without constraints on the
final values on their registers are also sometimes called cost-enriched
automata, weighted automata or counter automata, depending on your definitions
and side-constraints.

\contents{
  \item Automata intersection emptiness is PSPACE-complete
  \item Producing the formula is linear, but
  \item deciding a product of automata on a word is also PSPACE-hard
  \item try proving that finding a word accepted by a product AND satisfying additional constraints is IN PSPACE, e.g. that we are PSPACE-complete
}

@@

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute the Parikh image of products of Parikh automata.
\item A specialised method for quantifier elimination that blends well with \Calculus{} to efficiently compute quantifier-free Presburger representations of a given Parikh image.
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future studies.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\item Finally, a published description of the well-known bug in~\cite{generate-parikh-image} previously only described in folklore.
\end{itemize}

@@

To illustrate the key features of our approach, we start with two regular
languages, a commutative homomorphism $f$ on them that we care about, and a side
constraint for which we want to find values in $f$'s image over the intersection
of the wto languages. We will use the same two regular languages, recognised by
two automata; $\SomethingCSomething{}$ (Figure~\ref{fig:something-c-something})
and $\AcaOrBc{}$ (Figure~\ref{fig:aca-or-bc}) for both examples but change the
constraint and homomorphism $f$.

@@

Now assume that rather than the length (e.g. the count of the characters) we
want to find the whole Parikh image of the intersection of $\AcaOrBc{}$ and
$\SomethingCSomething{}$. In other words, now $f$ is a mapping to an object
(usually seen as a vector) counting how often each character occurs in a string. Since
extracting values from a vector is tiresome, we will use the shorthand notation
$\TransitionCount{a}$ to refer to the number of letters a that appear. E.g.
$\TransitionCount{a}(\text{curious and curiouser still}) = 1$. Note that we
still have the property that $\TransitionCount{o}(\text{homomorphism}) =
\TransitionCount{o}(\text{homo}) + \TransitionCount{o}(\text{morphism}) = 2 + 1
= 3$. If you treat the addition as element-wise vector addition, this approach
works for the vector version as well.

@@

\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \implies \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition = \FromLabelTo{\State}{}{}} - \sum_{\Transition = \FromLabelTo{}{}{\State}}
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \implies \StateVar_{\State}
} \\
& \AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \implies
  \left(\FinalStateVar_{\State} = 1 \text{ if } \State \in \AcceptingStates\right) \OrComp{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

@@

\contents{
  \item mention the bug here!
}

@@

\begin{equation}
  \begin{aligned}
  \ParikhMap(\Automaton):=\:
  &\LetterVar_{a} = \TransitionVar_{\FromLabelTo{0}{a}{1}} + \TransitionVar_{\FromLabelTo{1}{a}{3}} \land \LetterVar_{b} = \TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{2}{b}{1}} \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}} + \TransitionVar_{\FromLabelTo{2}{c}{3}}\\
  &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
  & \TransitionVar_{\FromLabelTo{0}{a}{1}} + \TransitionVar_{\FromLabelTo{0}{b}{2}} - 0 = 1 \land  \TransitionVar_{\FromLabelTo{1}{a}{3}} - \TransitionVar_{\FromLabelTo{2}{b}{1}} - \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0\\
  & \TransitionVar_{\FromLabelTo{2}{b}{1}} + \TransitionVar_{\FromLabelTo{2}{c}{3}} - \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0 \land \FinalStateVar_{3} + 0 - \TransitionVar_{\FromLabelTo{1}{a}{3}} - \TransitionVar_{\FromLabelTo{2}{c}{3}} = 0\\
  &\TransitionVar_{\FromLabelTo{0}{a}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{c}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{1}{a}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\TransitionVar_{\FromLabelTo{2}{b}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{2}{c}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{0}{a}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} \geq 1 \land \StateVar_{2} \geq 1\right)\\
  &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{1}{c}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{1} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{1}{a}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{2}{b}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{2} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{2}{c}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
  \end{aligned}
  \label{eq:parikh-formula}
  \end{equation}

  We see here that the formula encodes the choice of path at the start of the
  automaton ($\TransitionVar_{\FromLabelTo{0}{b}{2}} +
  \TransitionVar_{\FromLabelTo{0}{a}{1}} = 1$), and the dependency between that
  choice and the following outgoing transition, e.g. in
  $\TransitionVar_{\FromLabelTo{2}{d}{3}} -
  \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0$. Here we only have one choice of
  incoming flow; from $\FinalStateVar_3$, corresponding to the only accepting
  state. Simplifying the equation using standard algebra gives us this smaller
  set of clauses:

  \contents{\item FIXME: this needs rewriting for the new automaton}

  \begin{equation}
    \begin{aligned}
    &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
    & \LetterVar_{d} + \LetterVar_{a} = 1\\
    & \FinalStateVar_{3}  > \LetterVar_{d} \implies \StateVar_{1} > 0 \land \LetterVar_{d} > 0 \implies \StateVar_{2} > 0\\
    &\LetterVar_{c} > 0 \implies \StateVar_{1} > 0 \land \LetterVar_{e} > 0 \implies \StateVar_{2} > 0\\
    &\FinalStateVar_{3}  > \LetterVar_{d} \implies \StateVar_{3} > 0 \land \LetterVar_{d} > 0 \implies \StateVar_{3} > 0\\
    &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land (\FinalStateVar_{3}  - \LetterVar_{d}) \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \LetterVar_{d} \geq 1 \land \StateVar_{2} \geq 1\right)\\
    &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{3} + 1 \land (\FinalStateVar_{3}  - \LetterVar_{d}) \geq 1 \land \StateVar_{3} \geq 1\right)\\
    &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{3} + 1 \land \LetterVar_{d} \geq 1 \land \StateVar_{3} \geq 1\right)\\
    &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
    \end{aligned}
    \label{eq:parikh-formula:reduced}
    \end{equation}

Note that some of the or-clauses disappear since they contain false
inequalities of the type $x = x + 1$, arising from cycles. These clauses encode
the distance in a spanning tree from an accepting state, and the
unsatisfiability is by design, essentially ensuring that two states in a cycle
cannot vouch for each other's connectedness.

Assume we are really interested in occurrences with the letter e, i.e. we solve
for $\LetterVar_e > 0$. This implies that $\StateVar_{2} > 0$ (essentially, that
the state our transition of interest started in is reachable), in turn implying
that $\StateVar_{2} = \StateVar_{3} + 1 \land \LetterVar_{d} \geq 1 \land
\StateVar_{3} \geq 1$. Armed with this knowledge, we now know that $\StateVar_3
> 0$, which means that $\FinalStateVar_3 = 1$. Since we also know that
$\LetterVar_d \geq 1$, we know that $\LetterVar_a = 0$. This corresponds to
choosing the bottom path through the automaton. Applying this to the equation
and performing standard boolean logic propagation gives the following:

\begin{equation}
  \begin{aligned}
  &\LetterVar_{d} = 1 \land \LetterVar_{a} = 0 \\
  & \StateVar_{3} > -1 \\
  &\LetterVar_{c} = 0\\
  &\StateVar_{0} > 0 \implies \StateVar_{0} = \StateVar_{3} + 2\\
  \end{aligned}
  \label{eq:parikh-formula:choose-a}
  \end{equation}

  Since the final implication holds (we know $\StateVar_3 \geq 0$, and therefore
  $\StateVar_3 + 2 > 0$), we can disregard it. Thereby we have eliminated all
  quantified variables and are left with a formula containing only the free
  ones. Note that all constraints on $\LetterVar_e$ have disappeared after we
  asserted that it was strictly positive. This is no coincidence: since it is a
  loop, it can take any positive value.

  This example, in addition to showing how the encoding works, has hopefully
  also illustrated one of the key insights behind the design of \Calculus: the
  fact that even few additional constraints on the Parikh image can dramatically
  reduce the difficulty of computing it, as in this extreme case where only one
  additional choice in practice fixes the values of all other symbols in the image.

@@

To generalise the logic over Parikh images, we introdouce a morphism $\Map:\: \Sigma
\to \Monoid$ where $\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative monoid where the following holds:
\begin{itemize}
  \item $\Map(\epsilon) = 0_{\Monoid}$
  \item $\Map(a \Concat b) = \Map(a) +_{\Monoid} \Map(b)$, where $\Concat$ is
  the string concatenation operation
\end{itemize}

@@

We use the symbolic function $\FlowEq(\Automaton, \Filter)$ that generates a set
of existentially quantified linear inequalities with the following definition:
$$
\begin{aligned}
  & \FlowEq(\Automaton, \Filter) = \AndComp{\State \in \States}{\In(\State, \Filter) - \Out(\State, \Filter)}
\text{ $\geq 0$ if $\State \in \AcceptingStates$, $= 0$ otherwise}\\
  & \In(\State, \Filter) = \left(\text{$1$ if $\State = \InitialState$, otherwise $0$}\right) + \sum_{\Transition = \FromLabelTo{\State_0}{}{\State} \in \Transitions} \Filter(\Transition)\\
  & \Out(\State, \Filter) = \sum_{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions} \Filter(\Transition)
\end{aligned}
$$
Where we transparently assign fresh variables to each state $\State \in \States$
and existentially quantify them in the whole clause.

@@

>>>>

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{ebproof}
\usepackage{stackengine}

@@

It is possible to compute the Parikh image of any context-free language using
the method described in~\cite{generate-parikh-image}. Later improvements have
produced a construction taking at most linear time to
produce~\cite{muscholl-linear}. However, the resulting existentially quantified
clauses are costly to eliminate as the number of variables increases, in
practice making many real-world problems intractable.

In addition to the elimination of existentially quantified variables, another
problem arises for the common case of solving length constraints in string
solvers. Conjunctions of string constraints lead to the computation of Parikh
images of products (intersections) of regular languages represented as finite
automata. Applying the approach in~\cite{generate-parikh-image} would require
the eager computation of the product before its Parikh image. In several
instances we have observed while solving real-world string constraints, the
computation of the product of automata exhausts the memory of any machine
available due to the exponential blow-up in size of the product, quickly
becoming intractable as the number of automata in the product increases.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary Parikh-like
homomorphism over automata labels. This enables us to let both calculations
inform each other, eliminating unnecessary work, and pruning the size of the
partial products considered in the computation for a smaller memory footprint.
Moreover, the method can be used iteratively to tackle smaller chunks of the
product incrementally, thereby avoiding running out of memory at all. We
implement this calculus, along with the approximate method
of~\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker, in a tool that we call \Catra{}. \Catra{} allows solving constraints
over groups of synchronised (products of) Parikh automata, that is
(finite-alphabet) symbolic automata with transitions that increment and
decrement integer registers and that allows checking linear constraints on these
registers after running the automata. Parikh automata without constraints on the
final values on their registers are also sometimes called cost-enriched
automata, weighted automata or counter automata, depending on your definitions
and side-constraints. In the satisfaction mode, we are more formally solving the
intersection problem of Parikh automata (under homomorphism), known to be
PSPACE-complete~\cite{graph-queries}.

@@

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute (a homomorphism on) the
Parikh image of products of Parikh automata.
\item A specialised method for quantifier elimination that blends well with \Calculus{} to efficiently compute quantifier-free Presburger representations of a given Parikh image.
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future study.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\end{itemize}

@@

To illustrate the key features of our approach, we start with two regular
languages, a commutative homomorphism $f$ on them that we care about, and a side
constraint for which we want to find values in $f$'s image over the intersection
of the two languages. We will use the same two regular languages, recognised by
two automata; $\SomethingCSomething{}$ (Figure~\ref{fig:something-c-something})
and $\AcaOrBc{}$ (Figure~\ref{fig:aca-or-bc}) for both examples but change the
constraint and homomorphism $f$.

@@

Now assume that rather than the length (e.g. the count of the characters) we
want to find the whole Parikh image of the intersection of $\AcaOrBc{}$ and
$\SomethingCSomething{}$. In other words, now $f$ is a mapping to an object
(usually represented as a column vector) counting how often each character
occurs in a string. Since extracting values from a vector is tiresome, we will
use the shorthand notation $\TransitionCount{a}$ to refer to the number of
letters a that appear. E.g. $\TransitionCount{a}(\text{curious and curiouser
still}) = 1$. Note that we still have the property that
$\TransitionCount{o}(\text{homomorphism}) = \TransitionCount{o}(\text{homo}) +
\TransitionCount{o}(\text{morphism}) = 2 + 1 = 3$. If you treat the addition as
element-wise vector addition, this approach works for the vector version as
well.

@@

\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \implies \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition = \FromLabelTo{\State}{}{}} - \sum_{\Transition = \FromLabelTo{}{}{\State}}
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \implies \StateVar_{\State} > 0
} \\
& \AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \implies
  \left(\FinalStateVar_{\State} = 1 \text{ if } \State \in \AcceptingStates \land \right) \OrComp{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

@@

@@

\begin{equation}
  \begin{aligned}
  \ParikhMap(\Automaton):=\:
  &\LetterVar_{a} = \TransitionVar_{\FromLabelTo{0}{a}{1}} + \TransitionVar_{\FromLabelTo{1}{a}{3}} \land \LetterVar_{b} = \TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{2}{b}{0}} \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}} + \TransitionVar_{\FromLabelTo{2}{c}{3}}\\
  &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
  &\TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{0}{a}{1}} - \TransitionVar_{\FromLabelTo{2}{b}{0}} = 1 \land  \TransitionVar_{\FromLabelTo{1}{a}{3}} - \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0\\
  & \TransitionVar_{\FromLabelTo{2}{b}{0}} + \TransitionVar_{\FromLabelTo{2}{c}{3}} - \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0 \land \FinalStateVar_{3}  - \TransitionVar_{\FromLabelTo{1}{a}{3}} - \TransitionVar_{\FromLabelTo{2}{c}{3}} = 0\\
  &\TransitionVar_{\FromLabelTo{0}{a}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{c}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{1}{a}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\TransitionVar_{\FromLabelTo{2}{b}{0}} > 0 \implies \StateVar_{0} > 0 \land \TransitionVar_{\FromLabelTo{2}{c}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{0}{a}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} \geq 1 \land \StateVar_{2} \geq 1\right)\\
  &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{1}{c}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{1} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{1}{a}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{0} + 1 \land \TransitionVar_{\FromLabelTo{2}{b}{0}} \geq 1 \land \StateVar_{0} \geq 1\right) \lor \left(\StateVar_{2} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{2}{c}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
  \end{aligned}
  \label{eq:parikh-formula}
  \end{equation}

  We see here that the formula encodes the choice of path at the start of the
  automaton ($\TransitionVar_{\FromLabelTo{0}{b}{2}} +
  \TransitionVar_{\FromLabelTo{0}{a}{1}} -
  \TransitionVar_{\FromLabelTo{2}{b}{0}} = 1$), and the dependency between that
  choice and the following outgoing transition, e.g. in
  $\TransitionVar_{\FromLabelTo{1}{a}{3}} -
  \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0$. Here we only have one choice of
  incoming flow; from $\FinalStateVar_3$, corresponding to the only accepting
  state.

Note that one of the $\lor$-clauses can be immediately removed since it
contains the contradiction $\StateVar_{1} = \StateVar_{1} + 1$, arising from
cycles. These clauses encode the distance in a spanning tree from an accepting
state, and the unsatisfiability is by design, essentially ensuring that two
states in a cycle cannot vouch for each other's connectedness.

Assume we are really interested in occurrences with the letter a, i.e. we solve
for $\LetterVar_a > 0$. This implies that $\StateVar_{1} > 0 \land \StateVar_{3}
> 0$ (essentially, that the state our transition of interest started in is
reachable), in turn implying that $\FinalStateVar_3 = 1$. Plugging in these
facts, we get:

\begin{equation}
  \begin{aligned}
  &\LetterVar_{a} = 2 \land \LetterVar_{b} = 0 \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}}\\
  & \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \StateVar_{0} = \StateVar_{3} + 2\\
  \end{aligned}
  \label{eq:parikh-formula-calc}
  \end{equation}

We can then proceed to eliminate the remaining existentially quantified
variables to obtain the formula $\LetterVar_{a} = 2 \land \LetterVar_{b} = 0
\land \LetterVar_{c} \geq 0$ without too much effort.

  This example, in addition to showing how the encoding works, has hopefully
  also illustrated one of the key insights behind the design of \Calculus: the
  fact that even few additional constraints on the Parikh image can dramatically
  reduce the difficulty of computing it.

@@

To generalise the logic over Parikh images, we introduce a morphism $\Map:\: \Sigma
\to \Monoid$ where $\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative monoid where the following holds:
\begin{itemize}
  \item $\Map(\epsilon) = 0_{\Monoid}$
  \item $\Map(a \Concat b) = \Map(a) +_{\Monoid} \Map(b)$, where $\Concat$ is
  the string concatenation operation
\end{itemize}

@@

We use the symbolic function $\FlowEq(\Automaton, \Filter)$ that generates a set
of existentially quantified linear inequalities with the following definition:
$$
\begin{aligned}
  & \FlowEq(\Automaton, \Filter) = \sum\limits_{\State \in \AcceptingStates} \FinalStateVar_\State = 1 \land \AndComp{\State \in \States}{\In(\State, \Filter) - \Out(\State, \Filter)} = \Sink(\State) \\
  & \Sink(\State) = 0 \text{ if } \State \not\in \AcceptingStates, \FinalStateVar_\State \text{ otherwise.} \\
  & \In(\State, \Filter) = \StartFlow(\State) + \sum_{\Transition = \FromLabelTo{\State_0}{}{\State} \in \Transitions} \Filter(\Transition)\\
  & \StartFlow(\State)  = 1 \text{ if } \State = \InitialState, \text{ otherwise $0$.} \\
  & \Out(\State, \Filter) = \sum_{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions} \Filter(\Transition)
\end{aligned}
$$
Where we transparently assign fresh variables to each state $\State \in
\States$ as well as every accepting state $\State' \in \AcceptingStates$ and
existentially quantify them in the whole clause.

@@

Starting with~$\AcaOrBc{}$, where $\Map$ is the length function, in effect
$\Transition \mapsto 1$ for transitions, and the constraints that the length is
odd using the same trick as in \cref{sec:introduction:motivation}, we have the
definitions in \cref{fig:example:single:equivalences} (omitting existential
quantifiers and $x \geq 0$ for every variable to avoid clutter).

\begin{figure}[ht]
  We define $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$ as follows:

  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{start}{a}{sawA}) & = \TransitionVar_1 \\
        & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_2 \\
        & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{sawB}{b}{start}) & = \TransitionVar_4 \\
        & \Filter(\FromLabelTo{sawA}{a}{final}) & = \TransitionVar_5 \\
        & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
      \end{aligned}
    \end{equation*}    
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        % start
        &  \TransitionVar_4 = \TransitionVar_5 + \TransitionVar_2 - 1 \\
        % sawA
        & \TransitionVar_1 = \TransitionVar_5 \\
        % sawB
        & \TransitionVar_2 = \TransitionVar_4 + \TransitionVar_6 \\
        % final
        & \TransitionVar_5 + \TransitionVar_6 = \FinalStateVar_1 \\
        & \FinalStateVar_1 = 1 \\
      \end{aligned}  
    \end{equation*}    
  \end{minipage}
  \caption{Equivalences defining $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$
  respectively.}\label{fig:example:single:equivalences}
  \end{figure}

Propagating the equivalences from \FlowEq{} gives us (under the implicit
assumption that every RHS is $\geq 0$):
\begin{equation*}
  \begin{aligned}
    & \Filter(\FromLabelTo{start}{a}{sawA}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_4 + \TransitionVar_6 \\
    & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
    & \Filter(\FromLabelTo{sawB}{b}{start}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
    & \Filter(\FromLabelTo{sawA}{a}{final}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
  \end{aligned}
\end{equation*}
